Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-28 paper.tex
Finish first pass on sample cluster trees

\subsection{Tree inference from clusters of samples}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path with
complex patterns of shared mutations.
In this section we discuss the strategy used to infer likely tree
toplogies among the samples that share the same copying path.
In the majority of cases this copying path will consist of a single
node (no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The differences required to deal with
newly-detected putative recombinants are discussed below
in Section XXX.

With tens of thousands of samples being added to the ARG per day,
such clusters of sequences attaching to the same node
can be large, with complex patterns of shared mutations.
To avoid the implied large polytomy and unparsimonious
repetition of identical mutations among sibling nodes, we
infer a tree among these samples (the LS HMM has already

A natural
way to

We can think of the sequences attaching to a given node in the ARG
from a daily batch as a polytomy,
but with potentially shared mutations among the sibling nodes
(Figure~\ref{fig:overview_sc2ts}C).

@@

>>>>

\subsection{Tree inference from clusters of samples}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path, with
complex patterns of shared mutations among those clusters of samples.
In this section we discuss the strategy used to infer likely tree
topologies within these clusters of samples sharing the same copying path.
In the majority of cases the copying path will consist of a single
node (i.e., no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The special treatment afforded to
newly-detected putative recombinants are discussed below
in Section XXX.

With tens of thousands of samples being added to the ARG per day,
we can have clusters of hundreds of sequences attaching to the same node.
While some of these samples will require no extra mutations
(i.e, they are identical to the attachment node), in general there
will be complex patterns of shared mutations among the samples
reflecting their evolutionary relationships. A natural way to infer
these within-cluster evolutionary relationships is to use standard
tree-building algorithms.
We can infer a likely tree relating the
samples in a cluster independently of the other samples in a
daily batch, and then attach the tree (and mutations)
to the ARG at the node identified by the HMM.

We currently use the implementation of the UPGMA algorithm~\citep{Michener1957-tr}
from SciPy~\citep{Pauli2020-scipy} to build trees from sample
clusters, and map mutations back to this topology using maximum parsimony.
This approach was chosen mainly for simplicity and because of the
speed and reliability of the implementation.
An issue with the UPGMA algorithm is that it generates a strictly
binary tree, creating internal nodes
supported by no informative site (i.e., having no mutation immediately
ancestral to them). To avoid such false precision we post-process
to remove unsupported internal nodes, representing the relationship
between $k$ identical descendants of the node as a polytomy of size $k$.

There are well-known issues with using such a simple algorithm for inferring
evolutionary relationships~\citep{Felsenstein2004-inferring}.
Since this within-cluster tree building has a significant influence on the
overall ARG topology (see Results), implementing more sophisticated
tree building methods which keep track of the location of the required mutations
(rather then inferring post-hoc by parsimony), is a likely avenue
for improvements in overall inference quality.

@@

\subsection{Node dating}
[Discuss how we date nodes, and suggest caution on interpretation]
